From Net-Zero to Net-Negative: Assessing \hydrogen{} and \carbon{} Network Strategies in Europe

### 2023-11-23 report/submissions/journal/paper/paper.tex
review initial Results section

% make abbreviation for co2
\newcommand{\carbon}{CO$_2$}
\newcommand{\hydrogen}{H$_2$}
\newcommand{\hydrogengrid}{\hydrogen{}--Grid}
\newcommand{\carbongrid}{\carbon{}--Grid}
\newcommand{\scbase}{Baseline scenario}
\newcommand{\scCO}{CO$_2$~Grid scenario}
\newcommand{\scH}{H$_2$~Grid scenario}
\newcommand{\schybrid}{Hybrid scenario}

@@

\begin{figure}[ht!]
    \centering
    \includegraphics[width=\linewidth]{comparison/default/figures/90_nodes/cost_bar.png}
    \caption[short]{Total annual system cost for the European energy system for the different scenarios subdivided into groups of technologies. While the `Baseline' scenario has neither a \hydrogen{} or \carbon{} network, `Hybrid' is allowed to expand both. "Gas Infrastructure" combines gas facilities for power and heat production, "H$_2$ Infrastructure" combines H$_2$ production, transport and re-electrification. "Carbon Capture at Point Sources" combines all technologies with integrated carbon capture.}
    \label{fig:cost_bar}
\end{figure}

System cost differences between the scenarios are driven by renewable energy resources (solar, wind and bioenergy), the \carbon{} and \hydrogen{} networks, as well as carbon capture technologies (DAC and CC at point sources).
In the Baseline scenario, the absence of dedicated carbon and hydrogen transport networks requires the system to deploy carbon production technologies (DAC and CC at point sources) in the regions with carbon sequestration sites, and small amounts of hydrogen production technologies (mainly electrolysis) in each region to meet local hydrogen demand for industry and land transport fuel cells.
For CU and its respective carbon and hydrogen demands, the system chooses the sites where on-site production of carbon, hydrogen and the synthesis process is cheapest. This leads to renewable resources and storage solutions being deployed not necessarily in the regions with the best resources or lowest costs for clean electricity generation, but where they are situated closer to the energy demand locations of electrolysers and DAC.
Thus, total system costs in the baseline scenario are highest, while the scenarios with additional transport options for carbon and hydrogen achieve a more efficient system allocation and lower total system costs.

We can now closer examine the cost reductions and shifts in cost allocation when shifting from the baseline scenario to the other three network scenarios.
The implementation of a \carbon{} transport network (\textit{\carbon{}~Grid}) facilitates efficient point-source carbon capture, substantially reducing the dependency on DAC technologies. While DAC supplies around 50\% of the carbon needed in the Baseline scenario (400 Mt \carbon{}, see Fig.~\ref{fig:balance_captured_carbon}), in the \scCO scenario, DAC is only used to supply around 20\% of the carbon (150 Mt \carbon{}). The remaining carbon is supplied by additional carbon capture technology on all plants burning biomass in industry facilities and in CHPs as well as all industry facilities using gas (see Fig.~\ref{fig:captureshare_line}), because the cost to deploy on-site carbon capture technology is relatively cheaper than DAC. The \carbon{} transport network enables the transport of carbon from the locations of the cheapest captured carbon supply sources to the locations of the carbon demand sinks at CU amd CS sites, reducing the need for expensive DAC. The improved carbon management infrastructure also allows for a more strategic deployment of renewables, particularly in areas with high sequestration potential, leading to cost savings in renewable generation and storage.

@@

The \hydrogen{}~Grid scenario underscores the pivotal role of hydrogen infrastructure. The creation of a hydrogen transport network enables the cost-effective delivery of hydrogen from regions with a surplus of cheap renewable energy to high-demand industrial centers. This scenario shows a pronounced decrease in investments in solar and gas  infrastructure. The \hydrogen{} network reduces DAC similarly to the \carbon{} network, but still produces 200 Mt of carbon from DAC. CC technology on process emissions increases from 90\% to 100\%, on biomass CHPs from 60\% to 100\%, and on facilities using gas in industry from 50\% to 75\%. CC on biomass use in industry remains at 50\%, same as the \scbase. This variation highlights how the implementation of \hydrogen{} and \carbon{} networks unlocks specific carbon capture potentials.

While the cost-reduction in the Hybrid scenario compared to the \hydrogen~Grid scenario is marginal, the allocation of costs reveals a propagation of reduced DAC facilities and increased CC at point sources. The Hybrid scenario relies much heavier on investments into the \hydrogen{} network which outweigh investments in the \carbon{} network by a factor of 4.

@@

\begin{figure*}[h]
    \centering
    \begin{subfigure}{.5\textwidth}
        \centering
        \includegraphics[width=\linewidth]{full/figures/90_nodes/capacity_map_carbon.png}
        \caption{\carbon{} Sector.}
        \label{fig:capacity_map_carbon_co2}
    \end{subfigure}%
    \hfill
    \begin{subfigure}{.5\textwidth}
        \centering
        \includegraphics[width=\linewidth]{full/figures/90_nodes/capacity_map_hydrogen.png}
        \caption{\hydrogen{} Sector.}
        \label{fig:capacity_map_hydrogen_co2}
    \end{subfigure}
    \caption{Optimal production and transport capacities of the carbon and hydrogen sector in a net-zero energy system in Europe with both \carbon{} and \hydrogen{} network expansion (Hybrid).
    % Carbon network looks the same as in~\cite{morbeeOptimisedDeploymentEuropean2012}: two backbones, one in the nothern Europe other in south east.
    }
    \label{fig:capacity_maps}
\end{figure*}

@@

As depicted in Fig.~\ref{fig:capacity_maps}, the hybrid model incorporates the advantages from the a large hydrogen network and complements it with a carbon network. The carbon network unlocks cheap carbon in landlocked locations like  Poland, Bulgaria, and Austria and transports it to sequestration sites in the North Sea, Baltic Sea, Greece, and Italy, minimizing the need for DAC directly at the sequestration sites. At the same time, there is reduced need to transport hydrogen from Western regions like Spain and the UK to Central Europe.

A notable aspect of the hybrid model is the lack of overlap in the routes used by the hydrogen and carbon networks. The advantages of this model include a broader range of technologies contributing to its robustness, reduced reliance on DAC (which may be costlier than anticipated), and less land use for wind and solar due to the decreased necessity for DAC.

In summary, sequestration moves to sites with or close to the cheapest carbon sources. Adding the carbon network unlocks cheap inland carbon point sources to be transported to sequestration. FT fuel production moves to sites with or close to cheap hydrogen. The model prefers sites with locally available carbon from point sources and transport hydrogen from nearby cheap hydrogen producing locations, rather than transporting the carbon to the hydrogen site.

Furthermore, there are qualitative advantages to building out both a hydrogen and a carbon network that are not captured in the strict cost-benefit calculation employed in our optimization model. The resulting system may be more resilient towards unknown future developments in technologies, geopolitics, or policy goals.

% extend on flexibility gains

@@

The \hydrogengrid{} model emphasizes a substantial investment in hydrogen infrastructure, signifying a strategic bet on hydrogen's role in decarbonization. There is also a significant allocation for methanation and the expansion of electricity grids, underscoring the integration of hydrogen production with renewable energy sources.

% The Hybrid model, in contrast, adopts a more balanced investment approach, with notable allocations not just to hydrogen infrastructure but also to wind, solar, and CO2 infrastructure, as well as direct air capture (DAC) technology. This suggests a more comprehensive strategy that leverages a variety of technologies and energy sources to achieve emission reductions. The distributed investment in the Hybrid model indicates a systemic approach to decarbonization, integrating renewable generation, energy storage, carbon capture, and hydrogen production to address both energy and carbon management challenges.

\section{Limitations}
\label{sec:limitations}

@@

The technology costs used in the model rely on cost projections that incorporate reductions based on learning rates. These learning rates are derived from historical data, which may not necessarily be indicative of future trends. The model does not account for uncertainties arising from disruptive market behavior, such as the gas price peak that occurred in 2022.

@@

Lastly, the model's spatial and temporal resolution is insufficient to capture all relevant dynamics. Variations in renewable energy supply and energy demand below the applied time resolution, as well as more detailed interregional energy transport constraints, are currently not considered in the optimization. Transitioning to an hourly representation and a higher spatial resolution would increase the model's computational time and complexity, but would also enhance its robustness and validity.

In this study, we investigated the impact of a \carbon{} transport system with different carbon sequestration potentials on the optimal configuration of technologies in a fully sector-coupled energy system in Europe, focusing on the deployment of carbon capture and Utilization (CCU) and Carbon Capture and Storage (CCS) technologies. Our results indicate that increasing carbon sequestration potential leads to a decrease in total system costs, with a more pronounced cost reduction observed when a carbon transport network is present. Higher sequestration rates result in a shift from CCU and hydrogen infrastructure towards CCS and unabated fossil fuel usage compensated by carbon dioxide removal elsewhere in the system, particularly based on direct air and biomass carbon capture with sequestration.

Furthermore, we observed that with a \carbon{} transport network, the system tends to build out more carbon capture at point sources. With an increased sequestration potential, the share of technologies equipped with integrated carbon capture rises, particularly for Steam Methane Reforming (SMR) and Gas Combined Heat and Power (CHP) plants. Independent of a setup with or without \carbon{} transport, the spatial distribution of carbon capture technologies becomes more centralized, with installations concentrated near sequestration sites.

Despite limitations of the study considering fixed model assumptions and cost uncertainties, our findings contribute valuable insights into the role of a carbon transport system and carbon sequestration potential in the transition towards a climate-neutral and net-negative energy system. As Europe strives to achieve its 2050 climate neutrality target, understanding the interplay between carbon sequestration, CCU, and CCS technologies will be crucial for the design of effective policies and strategies.

@@

% \begin{figure*}
%     \centering
%     \includegraphics[width=\linewidth]{net-negative-0.1/full/figures/90_nodes/sankey_diagramm.png}
%     \caption{Sankey diagram of the optimal operation for a net-negative 10\% scenario.}
%     \label{fig:sankey_diagramm}
% \end{figure*}

% \begin{figure}
%     \centering
%     \includegraphics*[width=0.8\linewidth]{comparison/emission-reduction/figures/90_nodes/objective_heatmap.png}
%     \caption{Total annual system cost for the different scenarios and \carbon{} targets. While the `Baseline' scenario has neither a \hydrogen{} or \carbon{} network, `Hybrid' is allowed to expand both.}
%     \label{fig:objective_heatmap}
% \end{figure}

% \begin{figure}[h]
%     \centering
%     \includegraphics[width=\linewidth]{comparison/emission-reduction-full/figures/90_nodes/transmission_cost_bar.png}
%     \caption{Annual transmission system cost as a function of the net carbon removal scenarios considered in the study.}
%     \label{fig:transmission_cost_bar}
% \end{figure}

>>>>

% make abbreviation for co2
\newcommand{\carbon}{CO$_2$}
\newcommand{\hydrogen}{H$_2$}
\newcommand{\hydrogengrid}{\hydrogen{}--Grid}
\newcommand{\carbongrid}{\carbon{}--Grid}
\newcommand{\modBase}{Baseline model}
\newcommand{\modCO}{CO$_2$-Grid model}
\newcommand{\modH}{H$_2$-Grid model}
\newcommand{\modHybrid}{Hybrid model}

@@

\begin{figure}[ht!]
    \centering
    \includegraphics[width=\linewidth]{comparison/default/figures/90_nodes/cost_bar.png}
    \caption[short]{Total annual system cost for the European energy system for the different models subdivided into groups of technologies. While the `Baseline' model has neither a \hydrogen{} or \carbon{} network, `Hybrid' is allowed to expand both. "Gas Infrastructure" combines gas facilities for power and heat production, "H$_2$ Infrastructure" combines H$_2$ production, transport and re-electrification. "Carbon Capture at Point Sources" combines all technologies with integrated carbon capture.}
    \label{fig:cost_bar}
\end{figure}

System cost differences between the models are driven by renewable energy resources (solar, wind, and bioenergy), the \carbon{} and \hydrogen{} networks, as well as carbon capture technologies (DAC and CC at point sources).
In the Baseline model, the absence of dedicated carbon and hydrogen transport networks requires the system to deploy carbon production technologies (DAC and CC at point sources) in the regions with carbon sequestration sites, and small amounts of hydrogen production technologies (mainly electrolysis) in each region to meet local hydrogen demand for industry and land transport fuel cells.
For CU and its respective carbon and hydrogen demands, the system chooses the sites where on-site production of carbon, hydrogen and the synthesis process is cheapest. This leads to renewable resources and storage solutions being deployed not necessarily in the regions with the best resources or lowest costs for clean electricity generation, but where they are situated closer to the locations with carbon supply from CC at point sources, and to the locations of energy demand for electrolysers and DAC.
Thus, total system costs in the Baseline model are highest, while the models with additional transport options for carbon and hydrogen achieve a more efficient system allocation and lower total system costs.

We can now closer examine the cost reductions and shifts in cost allocation when shifting from the Baseline model to the other three network models.
The implementation of a \carbon{} transport network (\textit{\carbon{}~Grid}) facilitates efficient point-source carbon capture, substantially reducing the dependency on DAC technologies. While DAC supplies around 50\% of the carbon needed in the Baseline model (400 Mt \carbon{}, see Fig.~\ref{fig:balance_captured_carbon}), in the \modCO{}, DAC is only used to supply around 20\% of the carbon (150 Mt \carbon{}). The remaining carbon is supplied by additional carbon capture technology on all plants burning biomass in industry facilities and in CHPs as well as all industry facilities using gas (see Fig.~\ref{fig:captureshare_line}), because the cost to deploy on-site carbon capture technology is relatively cheaper than DAC. The \carbon{} transport network enables the transport of carbon from the locations of the cheapest captured carbon supply sources to the locations of the carbon demand sinks at CU amd CS sites, reducing the need for expensive DAC. The improved carbon management infrastructure also allows for a more strategic deployment of renewables, particularly in areas that combine high quality renewables and high sequestration potentials, leading to system-wide cost savings in renewable generation and storage.

@@

The \hydrogen{}-Grid model underscores the pivotal role of hydrogen infrastructure. The creation of a hydrogen transport network enables the cost-effective delivery of hydrogen from regions with a surplus of cheap renewable energy to high-demand industrial centers. This model shows a pronounced decrease in investments in solar generation and gas infrastructure. The \hydrogen{} network reduces DAC similarly to the \carbon{} network, but still produces 200 Mt of carbon from DAC. CC technology on process emissions increases from 90\% to 100\%, on biomass CHPs from 60\% to 100\%, and on facilities using gas in industry from 50\% to 75\%. CC on biomass use in industry remains at 50\%, same as the \modCO{} (see Figure~\ref{fig:captureshare_line}). This variation highlights how the implementation of \hydrogen{} and \carbon{} networks unlocks specific carbon capture potentials.

While the cost-reduction in the Hybrid model compared to the \hydrogen-Grid model is marginal, the allocation of costs reveals a propagation of reduced DAC facilities and increased CC at point sources. The Hybrid model relies slightly less on investments into the \hydrogen{} network than the \modH, but adds some investments into the \carbon{} network. Still, investments into the \hydrogen{} network outweigh investments into the \carbon{} network by a factor of four.

@@

@@

\begin{figure*}[h!]
    \centering
    \begin{subfigure}{.49\textwidth}
        \centering
        \includegraphics[width=\linewidth]{full/figures/90_nodes/capacity_map_hydrogen.png}
        \caption{\hydrogen{} Sector.}
        \label{fig:capacity_map_hydrogen_co2}
    \end{subfigure}
    \hfill
    \begin{subfigure}{.49\textwidth}
        \centering
        \includegraphics[width=\linewidth]{full/figures/90_nodes/capacity_map_carbon.png}
        \caption{\carbon{} Sector.}
        \label{fig:capacity_map_carbon_co2}
    \end{subfigure}%
    \caption{Optimal production and transport capacities of the carbon and hydrogen sector in a net-zero energy system in Europe with both \carbon{} and \hydrogen{} network expansion (Hybrid).
    % Carbon network looks the same as in~\cite{morbeeOptimisedDeploymentEuropean2012}: two backbones, one in the nothern Europe other in south east.
    }
    \label{fig:capacity_maps}
\end{figure*}

% increased BECCS, decreased DACS
% decreased methanation

The Hybrid model combines the advantages of both hydrogen and carbon networks. While the topology of the \hydrogen{}-grid roughly corresponds to that of the \hydrogen grid model, the \carbon{}-grid plays more of a complementary role. Fig.~\ref{fig:capacity_maps} shows the capacities of the two grids in the hybrid model, with the production and transport capacities given in kt/h. Note that despite similar overall capacities, the investment in the hydrogen infrastructure is four times the investment in the carbon infrastructure.

Similar to the \hydrogengrid model, the Hybrid model takes advantage of a large hydrogen network that enables hydrogen to be transported from centralized production sites in western regions such as Spain and the UK to decentralized FT production sites across the continent.
The carbon network supplements this with smaller clusters of networks. In addition to a larger carbon grid cluster in Central Europe, which transports carbon to the North Sea, the system is building three large linear carbon routes. These transport carbon from Romania and Bulgaria to Greece, from northern Italy to central Italy and from Hungary, the Czech Republic and Poland to the Baltic Sea. These enable the system to transport cheaply available carbon from inland biomass sources and sequester it (see Fig.~\ref{fig:captureshare_line}). This reduces the need for DAC in the system and the sequestration sites are shifted to the end points of the carbon networks. Second-order effects can be seen in the shift in FT production compared to the \hydrogengrid model. Regions that have both good renewable resources and sequestration potential could now forego sequestration and use local process emissions for FT production, as can be observed in Ireland. A more detailed of differences between the \hydrogengrid and the Hybrid model can be found in Fig.~.

A notable aspect of the hybrid model is the lack of overlap in the hydrogen and carbon network topologies which is tightly connected to the nodal carbon strategy. Depending on the exogeneous need in hydrogen for example from industry, and the costs for locally produced carbon, a region can either import hydrogen to produce FT fuels or export carbon.
\dots

Despite only small decrease in cost in comparison to the \hydrogengrid model, hybrid approach brings several advantages. These include a broader range of technologies contributing to its robustness, reduced reliance on DAC, which may be costlier than anticipated in the cost projections, and less land use for wind and solar due to the decreased necessity for DAC. The increase used of biomass with carbon capture, transport and sequestration leads to slight cost advantages in comparison to the \hydrogengrid model.

@@

Both systems show a stron increase in DAC

In this study\dots

\dots

@@

The technology costs used in the model rely on cost projections that incorporate reductions based on learning rates. These learning rates are derived from historical data, which may not necessarily be indicative of future trends. In particular, cost assumptions on DAC, electrolysis and \hydrogen{} pipeline costs, could have a strong impact in the model results.

The modelling results are heavily driven by the demand and emissions from industrial clusters. Allowing the model to relocate these and/or incorporate flexibility measures, may lead to less dependency on both \hydrogen{} and \carbon{} networks.

The technological flexibilities in the carbon sector might not be exploited to full extent. Therefore, the \carbon{} transport via truck or ship is not considered. Fischer-Tropsch facilities require to run on baseload with at least 90\% of their nominal capacity.

\dots

@@

@@

\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{net-negative-0.1/full/figures/90_nodes/sankey_diagramm.png}
    \caption{Sankey diagram of the optimal operation for a net-negative 10\% scenario.}
    \label{fig:sankey_diagramm}
\end{figure*}

\begin{figure}
    \centering
    \includegraphics*[width=0.8\linewidth]{comparison/emission-reduction/figures/90_nodes/objective_heatmap.png}
    \caption{Total annual system cost for the different scenarios and \carbon{} targets. While the `Baseline' scenario has neither a \hydrogen{} or \carbon{} network, `Hybrid' is allowed to expand both.}
    \label{fig:objective_heatmap}
\end{figure}

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{comparison/emission-reduction-full/figures/90_nodes/transmission_cost_bar.png}
    \caption{Annual transmission system cost as a function of the net carbon removal scenarios considered in the study.}
    \label{fig:transmission_cost_bar}
\end{figure}

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{difference/h2-only-full/figures/90_nodes/cost_map.png}
    \caption{Difference in cost investments between \hydrogengrid and Hybrid model. The left subfigure shows higher spendings per technology and region and transport system for the \hydrogengrid model, the right shows higher spendings in the Hydrid model.}
    \label{fig:cost_map_difference}
\end{figure}
